---
title: "Database Analysis"
author: "Gentry Miller"
format: 
  html:
    code-overflow: wrap
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: Contents
    df-print: kable
    theme: readable
execute: 
  freeze: auto
editor: visual
---

# An Investigation of Primate Life History Data Quality

It has been noted that primate life history data is sparse and often varies in reliability (Borries et al. 2013). The primary goals of this investigation are to evaluate and compare 5 databases containing primate life history statistics to assess their completeness and reliability while evaluating the efficacy of principal component analysis (PCA) as an advanced statistical method for identifying hidden trends in multivariate analysis.

## Setup

### Data Import & Cleaning

```{r}
#| label: data import script
#| code-fold: show

source("https://raw.githubusercontent.com/bygentry/LHDB-PCA/refs/heads/main/setup.R")
```

#### Filtering Primate Data

Four of the databases under investigation include data for more animals than primates - pulling out the primate data makes analysis cleaner and easier.

#### Missing Values and Point Correction

Our .csv of the Kappeler & Pereira database was transcribed by hand from a digital scan of the original paper publication. One value of adult female body mass (`BM.F`) in the Kappeler & Pereira data has a "?", making the object type 'character' instead of 'numeric' . Mixing number and non-number characters will result in forced NA for that value when converting to type numeric, so the '?' is removed to include the value in completeness tests. Additionally, several of the databases use -`999` as the "missing value" - these are converted to `NA` for all fields of type numeric to allow quantitative analysis.

#### Variable Selection

> See the [Variable_Comparisons page](https://docs.google.com/spreadsheets/d/12S0tICGxE4arILK3V4fUOHz7dU1SGTNLb7kZeripvZU/edit?gid=909701723#gid=909701723) for a quick side-by-side of how our variables are defined in each source. Individual database variable description pages can be accessed through the toolbar at the bottom of the linked page for more details.

\> Why did we choose these variables? Are they representative of LH characteristics?

```{r}
#| label: building dfs of key vars
#| code-fold: true

sources = c("Myrhvold et al. 2015", "de Magalhães & Costa 2009", "Ernest 2003", "Jones et al. 2009", "Kappeler & Pereira 2003")

## selecting vars for cross-database comparison
amniote_s <- amniote %>% 
  transmute(
    species = paste(genus, species),
    body_mass_u = adult_body_mass_g,
    body_mass_f = female_body_mass_g,
    body_mass_m = male_body_mass_g,
    AFR = female_maturity_d,
    gest_len = gestation_d,
    IBI = inter_litter_or_interbirth_interval_y,
    max_longevity = maximum_longevity_y,
    age_at_maturity_u = NA,
    age_at_maturity_f = female_maturity_d,
    age_at_maturity_m = male_maturity_d,
    weaning_age = weaning_d,
    source = sources[1]
  )

AnAge_s <- AnAge %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = `Body mass (g)`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = NA,
    gest_len = `Gestation/Incubation (days)`,
    IBI = `Inter-litter/Interbirth interval`,
    max_longevity = `Maximum longevity (yrs)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = `Female maturity (days)`,
    age_at_maturity_m = `Male maturity (days)`,
    weaning_age = `Weaning (days)`,
    source = sources[2]
  )

ernest_s <- ernest %>% 
  transmute(
    species = paste(Genus, species),
    body_mass_u = NA,
    body_mass_f = `mass(g)`,
    body_mass_m = NA,
    AFR = `AFR(mo)`,
    gest_len = `gestation(mo)`,
    IBI = `litters/year`,
    max_longevity = `max. life(mo)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `weaning(mo)`,
    source = sources[3]
  )

pan_s <- pan %>% 
  transmute(
    species = paste(MSW05_Genus, MSW05_Species),
    body_mass_u = `5-1_AdultBodyMass_g`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = `3-1_AgeatFirstBirth_d`,
    gest_len = `9-1_GestationLen_d`,
    IBI = `14-1_InterbirthInterval_d`,
    max_longevity = `17-1_MaxLongevity_m`,
    age_at_maturity_u = `23-1_SexualMaturityAge_d`,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `25-1_WeaningAge_d`,
    source = sources[4]
  )

kp_s <- kp %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = BM.U,
    body_mass_f = BM.F,
    body_mass_m = NA,
    AFR = AFR,
    gest_len = GL,
    IBI = IBI,
    max_longevity = NA,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = WA,
    source = sources[5]
  )
```

#### Standardizing Units

To make valid comparisons, variables across databases should be in common units. There are several changes to be made to standardize the units to meet these conditions:

1.  Longevity is reported in years

2.  All other time values are reported in days

3.  All mass values are reported in grams

Some of the time values are reported in months - I treat a month as the average length of a month in the Gregorian calendar in days and use these conversion factors throughout unit standardization:

$$ 1 \text{ year } = \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{365.25 \text{ days}}{12 \text{ months}} \approx \frac{30.44 \text{ days}}{1 \text{ month}} \\ $$

The following unit changes are necessary:

**Amniotes (Myrhvold et al. 2015):**

-   Inter-birth Interval (IBI) should be converted from years to days

**AnAge** **(de Magalhães & Costa 2009):**

-   No changes necessary

**Ernest (2003):**

-   Age at First Reproduction (AFR), Gestation Length (gest_len), and Weaning Age (weaning_age) should be converted from months to days

-   IBI should be converted from litters/year to days

    -   IBI is effectively days/birth (or days/litter) and can be derived from litters/year as such:

        $$ 1 \, \div \frac{x \text{ litters}}{1 \text{ year}} = \frac{x \text{ years}}{1 \text{ litter}} \,\times \, \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{x \text{ days}}{1\text{ litter}}
        $$

-   Maximum Longevity (max_longevity) should be converted from months to years

**PanTHERIA (Jones et al. 2009):**

-   Maximum Longevity should be converted from months to years

**Kappeler & Pereira (2003):**

-   IBI should be converted from months to days

```{r}
#| label: unit conversions
#| code-fold: true

days_per_year <- 365.25
months_per_year <- 12
days_per_month <- days_per_year/months_per_year

amniote_s <- amniote_s %>% 
  mutate(
    IBI = IBI * days_per_year
    # converts years to days
  )

ernest_s <- ernest_s %>% 
  mutate(
    AFR = AFR * days_per_month,
    # converts months to days
    gest_len = gest_len * days_per_month,
    # converts months to days
    IBI = days_per_year/IBI,
    # converts litters/year to IBI (days)
    max_longevity = max_longevity/months_per_year,
    # converts months to years
    weaning_age = weaning_age * days_per_month
    # converts months to days
  )

pan_s <- pan_s %>% 
  mutate(
    max_longevity = max_longevity/months_per_year
    # converts months to years
  )

kp_s <- kp_s %>% 
  mutate(
    IBI = IBI * days_per_month
    # converts months to days
  )
```

## Analysis

### Proportions of Missing Values

#### Pre-visualization Processing

```{r}
#| label: heat map prep
#| code-fold: true

# Variable % Missing Vals Heat Map

## combine values from key value dfs of each source into a single df "key_vals"
key_vals <- bind_rows(amniote_s, AnAge_s, ernest_s, kp_s, pan_s)

## this calculates the proportion of missing values for each life history trait
## and stores them in a "missigness" df that stores the trait, the source, 
## and the % missingness in 3 steps
## (1) pivots columns except for species and source, 
##     assigns variable names to trait and values to value 
## (2) groups traits sourcewise
## (3) calculates the proportion of NAs to valid entries for each trait
##     in each source and stores them alongside the source and trait name
missingness <- key_vals %>%
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")

missingness_no_sex <- key_vals %>%
  select(-any_of(c("body_mass_u", "body_mass_f", "body_mass_m", "age_at_maturity_u", "age_at_maturity_f", "age_at_maturity_m"))) %>% 
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")

## renames traits to look pretty on graph
missingness <- missingness %>%
  mutate(trait = case_when(
    trait == "max_longevity" ~ "Lifespan (Years)",
    trait == "IBI" ~ "Interbirth Interval (Days)",
    trait == "AFR" ~ "Age at First Reproduction (Days)",
    trait == "gest_len" ~ "Gestation Length (Days)",
    trait == "weaning_age" ~ "Weaning Age (Days)",
    trait == "body_mass_u" ~ "Adult Mass - Unspecified (g)",
    trait == "body_mass_m" ~ "Adult Mass - Male (g)",
    trait == "body_mass_f" ~ "Adult Mass - Female (g)",
    trait == "age_at_maturity_u" ~ "Sexual Maturity - Unspecified (Days)",
    trait == "age_at_maturity_m" ~ "Sexual Maturity - Male (Days)",
    trait == "age_at_maturity_f" ~ "Sexual Maturity - Female (Days)",
    TRUE ~ trait  # keep other traits unchanged
  ))

## renames traits to look pretty on graph
missingness_no_sex <- missingness_no_sex %>%
  mutate(trait = case_when(
    trait == "max_longevity" ~ "Lifespan (Years)",
    trait == "IBI" ~ "Interbirth Interval (Days)",
    trait == "AFR" ~ "Age at First Reproduction (Days)",
    trait == "gest_len" ~ "Gestation Length (Days)",
    trait == "weaning_age" ~ "Weaning Age (Days)",
    TRUE ~ trait  # keep other traits unchanged
  ))

## extracts the number of distinct species represented by each source
## then concatenates a label of the citation and number of primate species
dataset_labels <- key_vals %>%
  group_by(source) %>%
  summarize(n = n_distinct(species), .groups = "drop") %>%
  mutate(label = paste0(source, "\n(n = ", n, ")")) %>% 
  { setNames(.$label, .$source) }
```

```{r}
#| label: completeness bar chart prep
#| code-fold: true

## makes a new df using prop_missing values from 
## missingness df and averages them by source, then stores
## in new df mean_missing
mean_missing <- missingness %>% 
  group_by(source) %>% 
  summarise(
    mean_missingness = mean(prop_missing)
    )

## takes the inverse proportion of the mean missingness
## for each database to give mean completeness
mean_complete <- mean_missing %>% 
  mutate(mean_completeness = 1 - mean_missingness)

## removing sexed variables
mean_missing_no_sex <- missingness_no_sex %>% 
  group_by(source) %>% 
  summarise(
    mean_missingness_no_sex = mean(prop_missing)
  )

## takes the inverse proportion of the mean missingness
## for each database to give mean completeness
mean_complete_no_sex <- mean_missing_no_sex %>% 
  mutate(mean_completeness_no_sex = 1 - mean_missingness_no_sex)

composite_completeness <- mean_complete %>% 
  inner_join(mean_complete_no_sex, by = "source") %>% 
  mutate(
    cap = pmax(0, mean_completeness_no_sex - mean_completeness)
  )

composite_plot <- composite_completeness %>% 
  select(source, mean_completeness, cap) %>% 
  pivot_longer(
    cols = c(mean_completeness, cap),
    names_to = "segment",
    values_to = "value"
  ) %>% 
  mutate(
    segment = recode(segment,
                     mean_completeness = "All Key Vars",
                     cap = "Sexless Key Vars"
                     ),
    source = factor(source, levels = unique(source))
  )
```

#### Visualization & Interpretation

```{r}
#| label: heatmap visualization
#| code-fold: true
prop_missing <- ggplot(missingness, aes(x = source, y = trait, fill = prop_missing)) +
  geom_tile(color = "snow") +
  scale_fill_gradientn(
    colors = rev(colorspace::heat_hcl(100)),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(accuracy = 1),
    name = "% Missing"
    ) +
  scale_x_discrete(labels = dataset_labels) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Proportions of Missing Values Across Datasets",
    x = "Dataset", 
    y = "Trait")
```

Trends:

-   sex-reliant vars are most incomplete

-   sex-reliant vars for females are more complete than male or unknown

-   weaning age, IBI, and gestation length most complete across datasets (lifespan close 4th but K&P doesn't record longevity)

-   all datasets have at least one unreported variable of our key variable selection

```{r}
#| label: layered bar chart
#| code-fold: true

# make a named palette once so fill & color match perfectly
src_levels <- levels(composite_plot$source)
src_pal <- setNames(scales::hue_pal()(length(src_levels)), src_levels)

ggplot() +
  # draw Sexless first (white interior, colored outline)
  geom_col(
    data = composite_plot %>% filter(segment != "All Key Vars"),
    aes(x = source, y = value, color = source),
    fill = "white",
    width = 0.75,
    linewidth = 0.6,
    show.legend = FALSE   # <- key step: prevents outline-style legend
  ) +
  # draw All Key Vars on top (solid fill)
  geom_col(
    data = composite_plot %>% filter(segment == "All Key Vars"),
    aes(x = source, y = value, fill = source),
    width = 0.75
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_fill_manual(values = src_pal) +
  scale_color_manual(values = src_pal, guide = "none") +
  labs(
    x = "Source Database",
    y = "Mean Completeness",
    title = "Mean Completeness of Sexless vs All Key Vars"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 33, hjust = 1)
  )
```

```{r}
#| label: completeness bar plots
#| code-fold: true

p1 <- ggplot(mean_complete,
       aes(x = source, y = mean_completeness, fill = source)) +
  geom_col() +
  labs(
    title = "All Key Vars"
  ) +
  ylim(0, 1) + 
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )

p2 <- ggplot(mean_complete_no_sex,
       aes(x = source, y = mean_completeness_no_sex, fill = source)) +
  geom_col() +
  labs(
    title = "Sexless Key Vars"
  ) +
  ylim(0, 1) + 
  theme_bw() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )

ggarrange(p1, p2, 
          nrow = 1, 
          ncol = 2,
          legend = "bottom",
          common.legend = TRUE
          )
```

-   removing sexed vars improves completeness of all databases
-   check significance of difference\^ 2tail?

```{r}

```

## Sources

1.  Borries, C., Gordon, A. D., & Koenig, A. (2013). Beware of Primate Life History Data: A Plea for Data Standards and a Repository. PLOS ONE, 8(6), e67200. <https://doi.org/10.1371/journal.pone.0067200>

2.  Ernest, S.K.M. (2003), LIFE HISTORY CHARACTERISTICS OF PLACENTAL NONVOLANT MAMMALS. Ecology, 84: 3402-3402. <https://doi.org/10.1890/02-9002>

3.  de Magalhães JP, Costa J. A database of vertebrate longevity records and their relation to other life-history traits. J Evol Biol. 2009 Aug;22(8):1770-4. doi: 10.1111/j.1420-9101.2009.01783.x. Epub 2009 Jun 10. PMID: 19522730.

4.  de Magalhães, J. P., Abidi, Z., Dos Santos, G. A., Avelar, R. A., Barardo, D., Chatsirisupachai, K., Clark, P., De-Souza, E. A., Johnson, E. J., Lopes, I., Novoa, G., Senez, L., Talay, A., Thornton, D., To, P. K. P. (2024) "Human Ageing Genomic Resources: updates on key databases in ageing research." Nucleic Acids Research 52(D1):D900-D908.

5.  Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O’Dell, J., Orme, C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly, C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A., Price, S. A., Rigby, E. A., Rist, J., ... Purvis, A. (2009). PanTHERIA: A species-level database of life history, ecology, and geography of extant and recently extinct mammals. Ecology, 90(9), 2648–2648. <https://doi.org/10.1890/08-1494.1>

6.  Kappeler, P. M., Pereira, M. E., & Schaik, C. P. v. (2003). Primate life histories and socioecology. In P. M. Kappeler & M. E. Pereira (Eds.), *Primate life histories and socioecology* (pp. 1–20). The University of Chicago Press.

7.  Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. and Ernest, S.K.M. (2015), An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles. Ecology, 96: 3109-3109. <https://doi.org/10.1890/15-0846R.1>
