---
title: "Database Analysis"
author: "Gentry Miller"
format: 
  html:
    code-overflow: wrap
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: Contents
    df-print: kable
    theme: readable
editor: visual
---

# An Investigation of Primate Life History Data Quality

It has been noted that primate life history data is sparse and often varies in reliability (Borries et al. 2013). The primary goals of this investigation are to evaluate and compare 5 databases containing primate life history statistics to assess their completeness and reliability while evaluating the efficacy of principal component analysis (PCA) as an advanced statistical method for identifying hidden trends in multivariate analysis.

# Preliminaries

## Loading Packages and Data

### Loading Packages

```{r}
#| label: packages

options(repos = "http://cran.us.r-project.org")

pacman::p_load(psych, tidyverse, curl, mice, rlang, yaml, gridExtra, lme4, XML, lubridate, gratia, VIM)
```

### Importing Data

```{r}
#| label: data import
#| code-fold: true

# AMNIOTES 
amniote <- read.csv("https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/AmnioteDB2015/Amniote_Database_Aug_2015.csv")

# ANAGE
AnAge <- read_tsv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/AnAge/anage_data.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# ERNEST
ernest <- read_delim(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/Ernest/Mammal_lifehistories_v2.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# PANTHERIA
pan <- read_tsv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/PanTHERIA/PanTHERIA_1-0_WR05_Aug2008.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# Kappeler & Pereira 
kp <- read.csv(file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/Kappeler-Pereira/KP.csv")
```

### Filtering Primate Data

Four of the databases under investigation include data for more animals than primates - pulling out the primate data makes analysis cleaner and easier.

```{r}
#| label: primate data extraction
#| code-fold: true

## creating subset df for primates
amniote_p <- amniote %>% filter(order == "Primates")

AnAge_p <- AnAge %>% filter(Order == "Primates")

ernest_p <- ernest %>% filter(order == "Primates")

pan_p <- pan %>% filter(MSW05_Order == "Primates")

# Kappeler & Pereira is already only primates
```

## Data Wrangling & Analysis Prep

### Removing Irrelevant Variables

The Amniotes Database (Myrhvold et al. 2015) has six variables that apply only to reptiles and avians, e.g. incubation time. Since these have no values for mammals and are not represented in any other databases, they will unfairly negatively impact measurements of completeness for primate data. As such, I have decided to remove them to more accurately represent the completeness of *primate* life history statistics in the database.

```{r}
#| code-fold: true

## removes reptile and avian variables
amniote_p <- amniote_p %>% 
  select(-any_of(c(
    "egg_mass_g", "incubation_d", "fledging_age_d", "egg_width_mm", "egg_length_mm", "fledging_mass_g"
    )))
```

### Data Type Conversions and Missing Values

```{r}
#| code-fold: true

## converts -999 ('no data' value) to NA
amniote_p[amniote_p == -999] <- NA

ernest_p[ernest_p == -999] <- NA

pan_p[pan_p == -999] <- NA

## removes non-number characters (see Kappeler & Pereira Transcription Key page for details)
## not removing non-number characters will result in forced NA for value in row 195
kp <- kp %>% 
  mutate(
    BM.F = readr::parse_number(as.character(BM.F))
  )

## converting taxonomic vars to factors to aid in grouping
amniote_p <- amniote_p %>% 
  mutate(
    across(c(class, order, family, genus, species, subspecies, common_name), as.factor)
  )

AnAge_p <- AnAge_p %>% 
  mutate(
    across(c(Kingdom, Class, Order, Genus, Species), as.factor)
  )

ernest_p <- ernest_p %>% 
  mutate(
    across(c(order, family, Genus, species), as.factor)
  )

pan_p <- pan_p %>% 
  mutate(
    across(c(MSW05_Order, MSW05_Family, MSW05_Genus, MSW05_Species), as.factor)
  )

kp <- kp %>% 
  mutate(
    across(c(Family, Genus, Species), as.factor),
    BM.F = as.numeric(BM.F)
  )
```

### Extracting Key LH Variables

\> Why did we choose these variables? Are they representative?

```{r}
#| code-fold: true

## selecting vars for cross-database comparison
amniote_s <- amniote_p %>% 
  transmute(
    genus = genus,
    species = species,
    body_mass_u = adult_body_mass_g,
    body_mass_f = female_body_mass_g,
    body_mass_m = male_body_mass_g,
    AFR = female_maturity_d,
    gest_len = gestation_d,
    IBI = inter_litter_or_interbirth_interval_y,
    # convert to days by 365.25 * IBI
    max_longevity = maximum_longevity_y,
    age_at_maturity_u = NA,
    age_at_maturity_f = female_maturity_d,
    age_at_maturity_m = male_maturity_d,
    weaning_age = weaning_d,
  )

AnAge_s <- AnAge_p %>% 
  transmute(
    genus = Genus,
    species = Species,
    body_mass_u = `Body mass (g)`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = NA,
    gest_len = `Gestation/Incubation (days)`,
    IBI = `Inter-litter/Interbirth interval`,
    max_longevity = `Maximum longevity (yrs)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = `Female maturity (days)`,
    age_at_maturity_m = `Male maturity (days)`,
    weaning_age = `Weaning (days)`
  )

ernest_s <- ernest_p %>% 
  transmute(
    genus = Genus,
    species = species,
    body_mass_u = NA,
    body_mass_f = `mass(g)`,
    body_mass_m = NA,
    AFR = `AFR(mo)`,
    # convert AFR to days 30.438 * AFR
    gest_len = `gestation(mo)`,
    # convert gest_len to days 30.438 * gest_len
    IBI = `litters/year`,
    # convert to days by (365.25 * (1/'litters/year'))
    max_longevity = `max. life(mo)`,
    # convert to years max_longevity/12
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `weaning(mo)`
    # convert to days 30.438 * weaning_age
  )

pan_s <- pan_p %>% 
  transmute(
    genus = MSW05_Genus,
    species = MSW05_Species,
    body_mass_u = `5-1_AdultBodyMass_g`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = `3-1_AgeatFirstBirth_d`,
    gest_len = `9-1_GestationLen_d`,
    IBI = `14-1_InterbirthInterval_d`,
    max_longevity = `17-1_MaxLongevity_m`,
    # convert to years max_longevity/12
    age_at_maturity_u = `23-1_SexualMaturityAge_d`,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `25-1_WeaningAge_d`
  )

kp_s <- kp %>% 
  transmute(
    genus = Genus,
    species = Species,
    body_mass_u = BM.U,
    body_mass_f = BM.F,
    body_mass_m = NA,
    AFR = AFR,
    gest_len = GL,
    IBI = IBI,
    # convert to days by 30.438 * IBI
    max_longevity = NA,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = WA
  )
```

### Standardizing Units

To make valid comparisons, variables across databases should be in common units. There are several changes to be made to standardize the units to meet these conditions:

1.  Longevity is reported in years

2.  All other time values are reported in days

3.  All mass values are reported in grams

Therefore the following changes are necessary:

**Amniotes (Myrhvold et al. 2015):**

-   Inter-birth Interval (IBI) should be converted from years to days

**AnAge** **(de Magalhães & Costa et al. 2009):**

-   No changes necessary

**Ernest (2003):**

-   Age at First Reproduction (AFR), Gestation Length (gest_len), and Weaning Age (weaning_age) should be converted from months to days

-   IBI should be converted from litters/year to days

-   Maximum Longevity (max_longevity) should be converted from months to years

**PanTHERIA (Jones et al. 2009):**

-   Maximum Longevity should be converted from months to years

**Kappeler & Pereira (2003):**

-   IBI should be converted from months to days

To make these changes I will use these unit conversions:

$$
1 \text{ year } = \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{365.25 \text{ days}}{12 \text{ months}} = \frac{30.438 \text{ days}}{1 \text{ month}} \\
$$

```{r}
#| code-fold: true

amniote_s <- amniote_s %>% 
  mutate(
    IBI = IBI * 365.25
    # converts years to days
  )

ernest_s <- ernest_s %>% 
  mutate(
    AFR = AFR * 30.438,
    # converts months to days
    gest_len = gest_len * 30.438,
    # converts months to days
    IBI = (365.25 * (1/IBI)),
    # converts litters/year to IBI, then years to days
    max_longevity = max_longevity/12,
    # converts months to years
    weaning_age = weaning_age * 30.438
    # converts months to days
  )

pan_s <- pan_s %>% 
  mutate(
    max_longevity = max_longevity/12
    # converts months to years
  )

kp_s <- kp_s %>% 
  mutate(
    IBI = IBI * 30.438
    # converts months to days
  )
```

# Analysis

## Proportions of Missing Variables

### Preparation

#### Variable Selection

> See the [Variable_Comparisons page](https://docs.google.com/spreadsheets/d/12S0tICGxE4arILK3V4fUOHz7dU1SGTNLb7kZeripvZU/edit?gid=909701723#gid=909701723) for a quick side-by-side of how our variables are defined in each source. Individual database variable description pages can be accessed through the toolbar at the bottom of the linked page for more details.

```{r}
#| label: analysis

# Variable Heat Map
#
# task: extract variables of interest and combine values
#       from different datasets under a common name
# sherry made a holder "<name>_selected" df for the vars of interest
# then used bind_rows() to combine them, stored "source" and then
# created a new row in the composite dataframe for the "source" 
# variable (being the in text citation of the database publication)
# new_df <- 

```

```{r}
#| label: misc
vars <- colnames(pan_p)
count = 0
i = 0

```

# Sources

1.  Borries, C., Gordon, A. D., & Koenig, A. (2013). Beware of Primate Life History Data: A Plea for Data Standards and a Repository. PLOS ONE, 8(6), e67200. <https://doi.org/10.1371/journal.pone.0067200>

2.  Ernest, S.K.M. (2003), LIFE HISTORY CHARACTERISTICS OF PLACENTAL NONVOLANT MAMMALS. Ecology, 84: 3402-3402. <https://doi.org/10.1890/02-9002>

3.  de Magalhães JP, Costa J. A database of vertebrate longevity records and their relation to other life-history traits. J Evol Biol. 2009 Aug;22(8):1770-4. doi: 10.1111/j.1420-9101.2009.01783.x. Epub 2009 Jun 10. PMID: 19522730.

4.  de Magalhaes, J. P., Abidi, Z., Dos Santos, G. A., Avelar, R. A., Barardo, D., Chatsirisupachai, K., Clark, P., De-Souza, E. A., Johnson, E. J., Lopes, I., Novoa, G., Senez, L., Talay, A., Thornton, D., To, P. K. P. (2024) "Human Ageing Genomic Resources: updates on key databases in ageing research." Nucleic Acids Research 52(D1):D900-D908.

5.  Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O’Dell, J., Orme, C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly, C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A., Price, S. A., Rigby, E. A., Rist, J., ... Purvis, A. (2009). PanTHERIA: A species-level database of life history, ecology, and geography of extant and recently extinct mammals. Ecology, 90(9), 2648–2648. <https://doi.org/10.1890/08-1494.1>

6.  Kappeler, P. M., Pereira, M. E., & Schaik, C. P. v. (2003). Primate life histories and socioecology. In P. M. Kappeler & M. E. Pereira (Eds.), *Primate life histories and socioecology* (pp. 1–20). The University of Chicago Press.

7.  Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. and Ernest, S.K.M. (2015), An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles. Ecology, 96: 3109-3109. <https://doi.org/10.1890/15-0846R.1>
