---
title: "Database Analysis"
author: "Gentry Miller"
format: 
  html:
    code-overflow: wrap
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: Contents
    df-print: kable
    theme: readable
editor: visual
---

# An Investigation of Primate Life History Data Quality

It has been noted that primate life history data is sparse and often varies in reliability (Borries et al. 2013). The primary goals of this investigation are to evaluate and compare 5 databases containing primate life history statistics to assess their completeness and reliability while evaluating the efficacy of principal component analysis (PCA) as an advanced statistical method for identifying hidden trends in multivariate analysis.

## Preliminaries

### Loading Packages and Data

#### Loading Packages

```{r}
#| label: packages

options(repos = "https://cran.us.r-project.org")

pacman::p_load(
  psych, tidyverse, curl, mice, rlang, yaml, gridExtra, lme4, XML, lubridate, gratia, VIM
  )
```

#### Importing Data

```{r}
#| label: data import
#| code-fold: true

# AMNIOTES 
amniote <- readr::read_csv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/AmnioteDB2015/Amniote_Database_Aug_2015.csv",
  col_names = TRUE,
  show_col_types = FALSE
  )

# ANAGE
AnAge <- readr::read_tsv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/AnAge/anage_data.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# ERNEST
ernest <- readr::read_delim(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/Ernest/Mammal_lifehistories_v2.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# PANTHERIA
pan <- readr::read_tsv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/PanTHERIA/PanTHERIA_1-0_WR05_Aug2008.txt", 
  col_names = TRUE, 
  show_col_types = FALSE
  )

# Kappeler & Pereira 
kp <- readr::read_csv(
  file = "https://raw.githubusercontent.com/bygentry/LHDB-PCA/main/Data/Kappeler-Pereira/KP.csv",
  col_names = TRUE,
  show_col_types = FALSE
  )
```

#### Filtering Primate Data

Four of the databases under investigation include data for more animals than primates - pulling out the primate data makes analysis cleaner and easier.

```{r}
#| label: primate data extraction
#| code-fold: true

## creating subset df for primates
amniote_p <- amniote %>% dplyr::filter(order == "Primates")

AnAge_p <- AnAge %>% dplyr::filter(Order == "Primates")

ernest_p <- ernest %>% dplyr::filter(order == "Primates")

pan_p <- pan %>% dplyr::filter(MSW05_Order == "Primates")

# Kappeler & Pereira is already only primates
```

### Data Wrangling & Analysis Prep

> See the [Variable_Comparisons page](https://docs.google.com/spreadsheets/d/12S0tICGxE4arILK3V4fUOHz7dU1SGTNLb7kZeripvZU/edit?gid=909701723#gid=909701723) for a quick side-by-side of how our variables are defined in each source. Individual database variable description pages can be accessed through the toolbar at the bottom of the linked page for more details.

#### Removing Irrelevant Variables

The Amniotes Database (Myrhvold et al. 2015) has six variables that apply only to reptiles and avians, e.g. incubation time. Since these have no values for mammals and are not represented in any other databases, they will unfairly negatively impact measurements of completeness for primate data. As such, I have decided to remove them to more accurately represent the completeness of *primate* life history statistics in the database.

```{r}
#| label: removing reptile and avian vars
#| code-fold: true

## removes reptile and avian variables
amniote_p <- amniote_p %>% 
  select(-any_of(c(
    "egg_mass_g", "incubation_d", "fledging_age_d", "egg_width_mm", "egg_length_mm", "fledging_mass_g"
    )))
```

#### Missing Values and Point Correction

```{r}
#| label: converting missing vals to NA
#| code-fold: true

## converts -999 ('no data' value) to NA
missing.val.to.na <- function(df){
  df |> mutate(across(where(is.numeric), ~ na_if(.x, -999)))
}

amniote_p <- missing.val.to.na(amniote_p)
ernest_p <- missing.val.to.na(ernest_p)
pan_p <- missing.val.to.na(pan_p)

## Point Correction: one value of BM.F has a "?" in it, making the object type character instead of numeric (see Kappeler & Pereira Transcription Key for more details). 
## not removing non-number characters will result in forced NA for that value when converting to type numeric, so it must be corrected for
kp <- kp %>% 
  mutate(
    BM.F = readr::parse_number(as.character(BM.F)),
    BM.F = as.numeric(BM.F)
  )
```

#### Extracting Key LH Variables

\> Why did we choose these variables? Are they representative?

```{r}
#| label: building dfs of key vars
#| code-fold: true

sources = c("Myrhvold et al. 2015", "de Magalhães & Costa 2009", "Ernest 2003", "Jones et al. 2009", "Kappeler & Pereira 2003")

## selecting vars for cross-database comparison
amniote_s <- amniote_p %>% 
  transmute(
    species = paste(genus, species),
    body_mass_u = adult_body_mass_g,
    body_mass_f = female_body_mass_g,
    body_mass_m = male_body_mass_g,
    AFR = female_maturity_d,
    gest_len = gestation_d,
    IBI = inter_litter_or_interbirth_interval_y,
    max_longevity = maximum_longevity_y,
    age_at_maturity_u = NA,
    age_at_maturity_f = female_maturity_d,
    age_at_maturity_m = male_maturity_d,
    weaning_age = weaning_d,
    source = sources[1]
  )

AnAge_s <- AnAge_p %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = `Body mass (g)`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = NA,
    gest_len = `Gestation/Incubation (days)`,
    IBI = `Inter-litter/Interbirth interval`,
    max_longevity = `Maximum longevity (yrs)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = `Female maturity (days)`,
    age_at_maturity_m = `Male maturity (days)`,
    weaning_age = `Weaning (days)`,
    source = sources[2]
  )

ernest_s <- ernest_p %>% 
  transmute(
    species = paste(Genus, species),
    body_mass_u = NA,
    body_mass_f = `mass(g)`,
    body_mass_m = NA,
    AFR = `AFR(mo)`,
    gest_len = `gestation(mo)`,
    IBI = `litters/year`,
    max_longevity = `max. life(mo)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `weaning(mo)`,
    source = sources[3]
  )

pan_s <- pan_p %>% 
  transmute(
    species = paste(MSW05_Genus, MSW05_Species),
    body_mass_u = `5-1_AdultBodyMass_g`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = `3-1_AgeatFirstBirth_d`,
    gest_len = `9-1_GestationLen_d`,
    IBI = `14-1_InterbirthInterval_d`,
    max_longevity = `17-1_MaxLongevity_m`,
    age_at_maturity_u = `23-1_SexualMaturityAge_d`,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `25-1_WeaningAge_d`,
    source = sources[4]
  )

kp_s <- kp %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = BM.U,
    body_mass_f = BM.F,
    body_mass_m = NA,
    AFR = AFR,
    gest_len = GL,
    IBI = IBI,
    max_longevity = NA,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = WA,
    source = sources[5]
  )
```

#### Standardizing Units

To make valid comparisons, variables across databases should be in common units. There are several changes to be made to standardize the units to meet these conditions:

1.  Longevity is reported in years

2.  All other time values are reported in days

3.  All mass values are reported in grams

Some of the time values are reported in months - I treat a month as the average length of a month in the Gregorian calendar and use these conversion factors throughout unit standardization:

$$ 1 \text{ year } = \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{365.25 \text{ days}}{12 \text{ months}} \approx \frac{30.44 \text{ days}}{1 \text{ month}} \\ $$

The following changes are necessary:

**Amniotes (Myrhvold et al. 2015):**

-   Inter-birth Interval (IBI) should be converted from years to days

**AnAge** **(de Magalhães & Costa 2009):**

-   No changes necessary

**Ernest (2003):**

-   Age at First Reproduction (AFR), Gestation Length (gest_len), and Weaning Age (weaning_age) should be converted from months to days

-   IBI should be converted from litters/year to days

    -   IBI is effectively days/birth (or days/litter) and can be derived from litters/year as such:

        $$ 1 \, \div \frac{x \text{ litters}}{1 \text{ year}} = \frac{x \text{ years}}{1 \text{ litter}} \,\times \, \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{x \text{ days}}{1\text{ litter}}
        $$

-   Maximum Longevity (max_longevity) should be converted from months to years

**PanTHERIA (Jones et al. 2009):**

-   Maximum Longevity should be converted from months to years

**Kappeler & Pereira (2003):**

-   IBI should be converted from months to days

```{r}
#| label: unit conversions
#| code-fold: true

days_per_year <- 365.25
months_per_year <- 12
days_per_month <- days_per_year/months_per_year

amniote_s <- amniote_s %>% 
  mutate(
    IBI = IBI * days_per_year
    # converts years to days
  )

ernest_s <- ernest_s %>% 
  mutate(
    AFR = AFR * days_per_month,
    # converts months to days
    gest_len = gest_len * days_per_month,
    # converts months to days
    IBI = days_per_year/IBI,
    # converts litters/year to IBI (days)
    max_longevity = max_longevity/months_per_year,
    # converts months to years
    weaning_age = weaning_age * days_per_month
    # converts months to days
  )

pan_s <- pan_s %>% 
  mutate(
    max_longevity = max_longevity/months_per_year
    # converts months to years
  )

kp_s <- kp_s %>% 
  mutate(
    IBI = IBI * days_per_month
    # converts months to days
  )
```

## Analysis

### Proportions of Missing Variables

#### Pre-visualization Processing

```{r}
#| label: heat map prep
#| code-fold: true

# Variable Heat Map

## combine values from key value dfs of each source into a single df "key_vals"
key_vals <- bind_rows(amniote_s, AnAge_s, ernest_s, kp_s, pan_s)

## this calculates the proportion of missing values for each life history trait
## and stores them in a "missigness" df that stores the trait, the source, 
## and the % missingness in 3 steps
## (1) pivots columns except for species and source, 
##     assigns variable names to trait and values to value 
## (2) groups traits sourcewise
## (3) calculates the proportion of NAs to valid entries for each trait
##     in each source and stores them alongside the source and trait name
missingness <- key_vals %>%
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")

## renames traits to look pretty on graph
missingness <- missingness %>%
  mutate(trait = case_when(
    trait == "max_longevity" ~ "Lifespan (Years)",
    trait == "IBI" ~ "Interbirth Interval (Days)",
    trait == "AFR" ~ "Age at First Reproduction (Days)",
    trait == "gest_len" ~ "Gestation Length (Days)",
    trait == "weaning_age" ~ "Weaning Age (Days)",
    trait == "body_mass_u" ~ "Adult Mass - Unspecified (g)",
    trait == "body_mass_m" ~ "Adult Mass - Male (g)",
    trait == "body_mass_f" ~ "Adult Mass - Female (g)",
    trait == "age_at_maturity_u" ~ "Sexual Maturity - Unspecified (Days)",
    trait == "age_at_maturity_m" ~ "Sexual Maturity - Male (Days)",
    trait == "age_at_maturity_f" ~ "Sexual Maturity - Female (Days)",
    TRUE ~ trait  # keep other traits unchanged
  ))

## extracts the number of distinct species represented by each source
## then concatenates a label of the citation and number of primate species
dataset_labels <- key_vals %>%
  group_by(source) %>%
  summarize(n = n_distinct(species), .groups = "drop") %>%
  mutate(label = paste0(source, "\n(n = ", n, ")")) %>% 
  { setNames(.$label, .$source) }
```

#### Visualization

```{r}
#| label: heatmap visualization
#| code-fold: true
ggplot(missingness, aes(x = source, y = trait, fill = prop_missing)) +
  geom_tile(color = "snow") +
  scale_fill_gradientn(
    colors = rev(colorspace::heat_hcl(100)),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(accuracy = 1),
    name = "% Missing"
    ) +
  scale_x_discrete(labels = dataset_labels) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Missing Data Patterns Across Datasets", 
       x = "Dataset", 
       y = "Trait")
```

## Sources

1.  Borries, C., Gordon, A. D., & Koenig, A. (2013). Beware of Primate Life History Data: A Plea for Data Standards and a Repository. PLOS ONE, 8(6), e67200. <https://doi.org/10.1371/journal.pone.0067200>

2.  Ernest, S.K.M. (2003), LIFE HISTORY CHARACTERISTICS OF PLACENTAL NONVOLANT MAMMALS. Ecology, 84: 3402-3402. <https://doi.org/10.1890/02-9002>

3.  de Magalhães JP, Costa J. A database of vertebrate longevity records and their relation to other life-history traits. J Evol Biol. 2009 Aug;22(8):1770-4. doi: 10.1111/j.1420-9101.2009.01783.x. Epub 2009 Jun 10. PMID: 19522730.

4.  de Magalhães, J. P., Abidi, Z., Dos Santos, G. A., Avelar, R. A., Barardo, D., Chatsirisupachai, K., Clark, P., De-Souza, E. A., Johnson, E. J., Lopes, I., Novoa, G., Senez, L., Talay, A., Thornton, D., To, P. K. P. (2024) "Human Ageing Genomic Resources: updates on key databases in ageing research." Nucleic Acids Research 52(D1):D900-D908.

5.  Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O’Dell, J., Orme, C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly, C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A., Price, S. A., Rigby, E. A., Rist, J., ... Purvis, A. (2009). PanTHERIA: A species-level database of life history, ecology, and geography of extant and recently extinct mammals. Ecology, 90(9), 2648–2648. <https://doi.org/10.1890/08-1494.1>

6.  Kappeler, P. M., Pereira, M. E., & Schaik, C. P. v. (2003). Primate life histories and socioecology. In P. M. Kappeler & M. E. Pereira (Eds.), *Primate life histories and socioecology* (pp. 1–20). The University of Chicago Press.

7.  Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. and Ernest, S.K.M. (2015), An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles. Ecology, 96: 3109-3109. <https://doi.org/10.1890/15-0846R.1>
