---
title: "An Investigation of Primate Life History Data Quality"
author: "Gentry Miller"
format: 
  html:
    code-overflow: wrap
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: Contents
    df-print: kable
    theme: readable
execute: 
  freeze: auto
editor: visual
---

# Introduction

> “Given this general importance of life history data and with the scientific ideals of transparency and repeatability in mind, one would assume that researchers have long since agreed upon life history definitions and are using reliable and accurate data stored in standardized repositories. However, while various databases exist, standards for data quality and specification are lacking.” (Borries et al. 2013)

Life history theory has provided not just a framework to understand how organisms maximize fitness, but a lexicon of variables that represent a shared currency across organisms. Key variables have emerged to characterize growth (weaning age ), reproduction (age at sexual maturity, first reproduction), and duration of reproduction (lifespan). Despite the widespread use of such measures, researchers face consistent methodological challenges in collating life history data for comparative analyses including access to a substantial range, differences in definitions, discrepancies in data quality, and decision trees on which variables to use and how. 

Here we present a guide that addresses these and other issues for primatologists who want to employ life history data. We review the characteristics of five databases that contain life history data on primates. We also present a frequently advocated approach to summarize variation across traits using dimension reduction analysis, with which we summarize data here. 

The databases assessed here are the PanTHERIA (Jones et al. 2009), AnAge (de Magalhães & Costa 2009), Amniotes (Myrhvold et al. 2015), Ernest (2003), and Kappeler & Pereira (2003) databases on life history statistics.

# Database Analysis

## Preliminaries

### Data Cleaning

```{r}
#| label: data import script
#| echo: false

source("https://raw.githubusercontent.com/bygentry/LHDB-PCA/refs/heads/main/setup.R")
```

#### Filtering Primate Data

Four of the databases under investigation (all but Kappeler & Pereira (2003)) include data for more animals than primates - our focus is on the quality of *primate* life history data, so all other taxa were excluded from our tests.

#### Missing Values and Point Correction

Our .csv of the Kappeler & Pereira database was transcribed by hand from a digital scan of the original paper publication. One value of adult female body mass (`BM.F`) in the Kappeler & Pereira data has a "?", making the object type 'character' instead of 'numeric' . Mixing number and non-number characters would result in a forced NA for that value when converting to type numeric, so the '?' is removed to include the value in completeness tests despite its uncertainty. Additionally, several of the databases use -`999` as the "missing value" - these are converted to `NA` for all fields of type numeric to allow quantitative analysis.

#### Variable Definitions and Selection

Each database records a number of life history statistics, any of which are reported in several of the databases as they are core life history traits (i.e. inter-birth interval, gestation length, etc). See how the variables are defined in each source database [here](https://bygentry.github.io/LHDB-PCA/Variable-Definitions.html). Rather than evaluating the completeness and reliability of every variable across all databases, we focus on a subset of “key life history traits” that characterize primate reproductive life history and are present in all or most of the databases examined. These traits serve as a representative sample of variables commonly used in taxonomic comparisons in life history research:

| Trait                                |
|:-------------------------------------|
| Adult Body Mass (Female)             |
| Adult Body Mass (Male)               |
| Adult Body Mass (Unspecified)        |
| Age at First Reproduction            |
| Age at Sexual Maturity (Female)      |
| Age at Sexual Maturity (Male)        |
| Age at Sexual Maturity (Unspecified) |
| Gestation Length                     |
| Inter-birth Interval                 |
| Lifespan                             |
| Weaning Age                          |

: **Table 1.** "Key Life History Traits" selected for cross-database comparison

```{r}
#| label: building dfs of key vars
#| code-fold: true

sources = as.factor(c("Myrhvold et al. 2015", "de Magalhães & Costa 2009", "Ernest 2003", "Jones et al. 2009", "Kappeler & Pereira 2003"))

## selecting vars for cross-database comparison
amniote_s <- amniote %>% 
  transmute(
    species = paste(genus, species),
    body_mass_u = adult_body_mass_g,
    body_mass_f = female_body_mass_g,
    body_mass_m = male_body_mass_g,
    AFR = female_maturity_d,
    gest_len = gestation_d,
    IBI = inter_litter_or_interbirth_interval_y,
    max_longevity = maximum_longevity_y,
    age_at_maturity_u = NA,
    age_at_maturity_f = female_maturity_d,
    age_at_maturity_m = male_maturity_d,
    weaning_age = weaning_d,
    source = sources[1]
  )

AnAge_s <- AnAge %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = `Body mass (g)`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = NA,
    gest_len = `Gestation/Incubation (days)`,
    IBI = `Inter-litter/Interbirth interval`,
    max_longevity = `Maximum longevity (yrs)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = `Female maturity (days)`,
    age_at_maturity_m = `Male maturity (days)`,
    weaning_age = `Weaning (days)`,
    source = sources[2]
  )

ernest_s <- ernest %>% 
  transmute(
    species = paste(Genus, species),
    body_mass_u = NA,
    body_mass_f = `mass(g)`,
    body_mass_m = NA,
    AFR = `AFR(mo)`,
    gest_len = `gestation(mo)`,
    IBI = `litters/year`,
    max_longevity = `max. life(mo)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `weaning(mo)`,
    source = sources[3]
  )

pan_s <- pan %>% 
  transmute(
    species = paste(MSW05_Genus, MSW05_Species),
    body_mass_u = `5-1_AdultBodyMass_g`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = `3-1_AgeatFirstBirth_d`,
    gest_len = `9-1_GestationLen_d`,
    IBI = `14-1_InterbirthInterval_d`,
    max_longevity = `17-1_MaxLongevity_m`,
    age_at_maturity_u = `23-1_SexualMaturityAge_d`,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `25-1_WeaningAge_d`,
    source = sources[4]
  )

kp_s <- kp %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = BM.U,
    body_mass_f = BM.F,
    body_mass_m = NA,
    AFR = AFR,
    gest_len = GL,
    IBI = IBI,
    max_longevity = NA,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = WA,
    source = sources[5]
  )
```

#### Standardizing Units

To make valid comparisons, variables across databases should be in common units. There are several changes to be made to standardize the units to meet these conditions:

1.  Longevity is reported in years

2.  All other time values are reported in days

3.  All mass values are reported in grams

Some of the time values are reported in months - we treat a month as the average length of a month in the Gregorian calendar in days and use these conversion factors throughout unit standardization:

$$ 1 \text{ year } = \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{365.25 \text{ days}}{12 \text{ months}} \approx \frac{30.44 \text{ days}}{1 \text{ month}} \\ $$

The following unit changes are necessary:

**Amniotes (Myrhvold et al. 2015):**

-   Inter-birth Interval (IBI) should be converted from years to days

**AnAge** **(de Magalhães & Costa 2009):**

-   No changes necessary

**Ernest (2003):**

-   Age at First Reproduction (AFR), Gestation Length (gest_len), and Weaning Age (weaning_age) should be converted from months to days

-   IBI should be derived from litters/year

    -   IBI is effectively days/birth (or days/litter) and can be derived from litters/year as such:

        $$ 1 \, \div \frac{x \text{ litters}}{1 \text{ year}} = \frac{x \text{ years}}{1 \text{ litter}} \,\times \, \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{x \text{ days}}{1\text{ litter}}
        $$

-   Maximum Longevity (max_longevity) should be converted from months to years

**PanTHERIA (Jones et al. 2009):**

-   Maximum Longevity should be converted from months to years

**Kappeler & Pereira (2003):**

-   IBI should be converted from months to days

```{r}
#| label: unit conversions
#| code-fold: true

days_per_year <- 365.25
months_per_year <- 12
days_per_month <- days_per_year/months_per_year

amniote_s <- amniote_s %>% 
  mutate(
    IBI = IBI * days_per_year
    # converts years to days
  )

ernest_s <- ernest_s %>% 
  mutate(
    AFR = AFR * days_per_month,
    # converts months to days
    gest_len = gest_len * days_per_month,
    # converts months to days
    IBI = days_per_year/IBI,
    # converts litters/year to IBI (days)
    max_longevity = max_longevity/months_per_year,
    # converts months to years
    weaning_age = weaning_age * days_per_month
    # converts months to days
  )

pan_s <- pan_s %>% 
  mutate(
    max_longevity = max_longevity/months_per_year
    # converts months to years
  )

kp_s <- kp_s %>% 
  mutate(
    IBI = IBI * days_per_month
    # converts months to days
  )
```

### Pre-vis Processing & Stats Test Setup

```{r}
#| label: heat map prep
#| code-fold: true
# Variable % Missing Vals Heat Map Prep

## combine values from key value dfs of each source into a single df "key_vals"
key_vals <- bind_rows(amniote_s, AnAge_s, ernest_s, kp_s, pan_s)

key_long <- key_vals %>%
  pivot_longer(
    cols = -c(species, source),
    names_to = "trait_raw",
    values_to = "value"
  )

trait_labels <- c(
  max_longevity = "Lifespan (Years)",
  IBI = "Interbirth Interval (Days)",
  AFR = "Age at First Reproduction (Days)",
  gest_len = "Gestation Length (Days)",
  weaning_age = "Weaning Age (Days)",
  body_mass_u = "Adult Mass - Unspecified (g)",
  body_mass_m = "Adult Mass - Male (g)",
  body_mass_f = "Adult Mass - Female (g)",
  age_at_maturity_u = "Sexual Maturity - Unspecified (Days)",
  age_at_maturity_m = "Sexual Maturity - Male (Days)",
  age_at_maturity_f = "Sexual Maturity - Female (Days)"
)

scope_df <- key_long %>%
  group_by(source, trait_raw) %>%
  summarise(
    n_total = n(),
    n_nonmissing = sum(!is.na(value)),
    prop_missing_panel = mean(is.na(value)),
    .groups = "drop"
  ) %>%
  mutate(
    in_scope = (n_nonmissing / n_total) >= 0.01,
    prop_missing_within_scope = if_else(in_scope, prop_missing_panel, NA_real_),
    trait = recode(trait_raw, !!!trait_labels)
  )

scope_summary <- scope_df %>%
  group_by(source) %>%
  summarise(
    scope_coverage = mean(in_scope),                              
    panel_completeness = 1 - mean(prop_missing_panel),            
    within_scope_completeness = 1 - mean(prop_missing_within_scope, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: legacy missingness dataframe
#| echo: false
#| eval: false  

# from old version of code, keeping just in case it becomes useful again

## this calculates the proportion of missing values for each life history trait
## and stores them in a "missigness" df that stores the trait, the source, 
## and the % missingness in 3 steps
## (1) pivots columns except for species and source, 
##     assigns variable names to trait and values to value 
## (2) groups traits sourcewise
## (3) calculates the proportion of NAs to valid entries for each trait
##     in each source and stores them alongside the source and trait name
missingness <- key_vals %>%
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")

missingness <- missingness %>%
  mutate(trait = dplyr::recode(trait, !!!trait_labels, .default = trait))

## extracts the number of distinct species represented by each source
## then concatenates a label of the citation and number of primate species
dataset_labels <- key_vals %>%
  group_by(source) %>%
  summarize(n = n_distinct(species), .groups = "drop") %>%
  mutate(label = paste0(source, "\n(n = ", n, ")")) %>% 
  { setNames(.$label, .$source) }
```

```{r}
#| label: permutation test
#| code-fold: true
# One-tailed Permutation Test

# create list of sex dependent variable names for checking sex differentiation
sex_dep_vars <- c("Sexual Maturity - Female (Days)", "Sexual Maturity - Male (Days)", "Sexual Maturity - Unspecified (Days)", "Adult Mass - Female (g)", "Adult Mass - Male (g)", "Adult Mass - Unspecified (g)")

# create a new df for the permutation test
## completeness - the inverse proportion of missing values
## sexed - a boolean representing a trait is stratified by sex
perm_df <- scope_df %>% 
  filter(in_scope) %>% 
  mutate(
    completeness = 1 - prop_missing_panel,        # inverse proportion of missingness for each trait
    sexed = trait %in% sex_dep_vars         # boolean: TRUE if variable is sex stratified, else FALSE 
  )

# creates a function for the permutation test
perm_test <- function(df, n = 10000, seed = 1){
  set.seed(seed)

  x <- df$completeness
  sexed <- df$sexed

  n_sexed <- sum(sexed)
  n_non <- sum(!sexed)

  if(n_sexed == 0 || n_non == 0){
    return(tibble(
      mean_sexed = NA_real_,
      mean_non = NA_real_,
      obs_diff = NA_real_,
      p_value = NA_real_,
      n_sexed = n_sexed,
      n_non = n_non
    ))
  }

  mu_sexed <- mean(x[sexed])
  mu_non <- mean(x[!sexed])
  obs <- mu_non - mu_sexed

  perm_stats <- replicate(n, {
    perm <- sample(sexed)
    mean(x[!perm]) - mean(x[perm])
  })

  tibble(
    mean_sexed = mu_sexed,
    mean_non = mu_non,
    obs_diff = obs,
    p_value = mean(perm_stats >= obs), 
    n_sexed = n_sexed,
    n_non = n_non
  )
}
```

## Proportions of Missing Values

One of the primary issues with primate life history data is how few species are represented. None of the databases inspected in this study represent more than 70% of primate species, highlighting the ever-present need for more data.

| Database | N species | Proportion of Primates Represented\* |
|----|----|----|
| Amniotes (Myhrvold et al. 2015) | 365 | .68 |
| AnAge (de Magalhães & Costa 2009) | 174 | .33 |
| Ernest (2003) | 156 | .29 |
| PanTHERIA (Jones et al. 2009) | 376 | .70 |
| Kappeler & Pereira (2003) | 310 | .58 |

: **Table 2.** Proportions of primate species represented by each database. \*Calculated as N/535, the number of species recognized by the IUCN / SSC Primate Specialist Group (2023)

Of the species we have data for, many are lacking at least one important life history statistic. Heat map analysis of the proportion of missing values to data points for each key life history trait reveal several patterns in the short-comings of the data we have available:

#### Visualization & Interpretation

```{r}
#| label: heatmap visualization 1
#| fig-cap: "**Figure 1.** Heat map of missing values for key life history traits across all five databases. Low heat (yellow) signifies few missing values while high heat (red) signifies many missing values. Grey indicates that the trait was not recorded in a particular source."
#| echo: false

ggplot(scope_df, aes(x = source, y = trait)) +
  geom_tile(aes(fill = prop_missing_within_scope), color = "snow") +
  geom_tile(
    data = dplyr::filter(scope_df, !in_scope),
    fill = "grey85",
    color = "snow"
  ) +
  scale_fill_gradientn(
    colors = rev(colorspace::heat_hcl(100)),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(accuracy = 1),
    name = "% Missing\n(within scope)",
    na.value = "grey85"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Proportions of Missing Values for Key Traits",
    x = "Dataset",
    y = "Trait"
  )

```

notes:

-   recall: Ernest's IBI is derived from litters/year

<!-- -->

-   sex-stratified vars are most unreported

-   age at maturity vars for females are more complete than male when both sexes are reported

-   weaning age, IBI, and gestation length are the most complete across all data sets (lifespan is a close 4th but K&P don't report any values for it - same with AnAge and AFR )

-   **all databases have at least one unreported variable of our key variable selection**

Of our key life history traits, most databases only record data for around 60% of them (mostly just **not** the sex-stratified versions of body mass and age at maturity). When testing the mean completeness of the databases using all of the key traits versus only the ones they record, their completeness scores obviously improved. This may appear to call into question our selection of traits, but really the trait scope is under represented because the missing traits are the same measurement for different sex classes.

```{r}
scope_summary %>%
  mutate(
    scope_coverage = scales::percent(scope_coverage, accuracy = 1),
    panel_completeness = round(panel_completeness, 3),
    within_scope_completeness = round(within_scope_completeness, 3)
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    source = "Database",
    scope_coverage = "Trait scope coverage",
    within_scope_completeness = "Completeness (within scope)",
    panel_completeness = "Completeness (full panel)"
  ) %>%
  gt::tab_caption(gt::md("**Table 3.** Scope coverage vs completeness (panel vs within-scope)."))
```

I ran a one-tailed permutation test to compare the mean completeness of the sex-stratified and non-sex-stratified variables to determine of sex-stratified traits are systematically less complete than non-stratified as a test of whether the stratified variables truly have worse completeness or if they are just under reported.

-   began re-applying permutation test to sexed vs non sexed variable completeness
    -   this informs us about the completeness of sex stratified traits compared to non stratified and informs us if sex stratified traits are systematically less complete
    -   if sex stratified traits are systematically less complete it may suggest that they are inherently harder to compile data on (expected)

```{r}
#| label: permutation test results
#| echo: false
perm_res <- perm_df %>%
  group_by(source) %>%
  group_modify(~ perm_test(.x)) %>%
  ungroup()

perm_res %>%
  mutate(
    mean_sexed = round(mean_sexed, 3),
    mean_non = round(mean_non, 3),
    obs_diff = round(obs_diff, 3),
    p_value = signif(p_value, 3)
  ) %>%
  gt() %>%
  cols_label(
    source = "Database",
    mean_non = "Non-sex-stratified traits",
    mean_sexed = "Sex-stratified traits",
    obs_diff = "Δ completeness (non − sexed)",
    p_value = "p (one-tailed)",
    n_non = "k (non)",
    n_sexed = "k (sexed)"
  ) %>%
  tab_spanner(
    label = "Mean completeness (within scope)",
    columns = c(mean_non, mean_sexed)
  ) %>%
  tab_caption(
    md("**Table 4.** Scope-aware permutation test comparing completeness of non-sex-stratified vs sex-stratified traits within each database’s in-scope trait set (n = 10,000 label permutations). One-tailed p tests whether non-sex-stratified traits are more complete.")
  )

```

When completeness is evaluated only among traits that are in the scope of each database, sex-stratified variables are not uniformly less complete. In fact, for Ernest (2003), PanTHERIA (Jones et al. 2009), and Kappeler & Pereira (2003), sex-stratified traits show higher mean completeness than non-sex-stratified traits, reflected by negative observed differences. In contrast, AnAge and the Amniote database show higher completeness for non-sex-stratified traits, although differences are not statistically significant. These results suggest that the apparent incompleteness of sex-stratified variables observed in panel-based analyses is largely driven by scope differences rather than systematic data loss within sex-stratified traits.

```{r}
#| label: layered bar chart
#| fig-cap: "**Figure 2.** Trait scope coverage vs within scope completeness across primate life history databases. Scope coverage represents the proportion of our select key traits recorded by a given source and the within scope completeness represents the mean completeness calculated only for traits recorded by that database."
#| echo: false

scope_long <- scope_summary %>% 
  pivot_longer(
    cols = c(scope_coverage, within_scope_completeness),
    names_to = "metric",
    values_to = "value"
  )

ggplot(scope_long, aes(x = source, y = value, fill = metric)) +
  geom_col(position = "dodge", width = .7) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_fill_manual(values = c(scope_coverage = "orange3", within_scope_completeness = "navyblue"),
  labels = c(scope_coverage = "Trait scope coverage", within_scope_completeness = "Completeness (within scope)"
  )) +
  labs(
    x = "Database",
    y = "Proportion",
    title = "Trait scope vs completeness across primate life history databases"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 33, hjust = 1)
  )
```

When trait scope is taken into account, the apparent incompleteness of sex-stratified variables across databases is largely explained by differences in database design. Databases such as Ernest (2003), PanTHERIA (Jones et al. 2009), and Kappeler & Pereira (2003) do not uniformly report sex-specific measures of adult body mass or age at maturity, which inflates missingness when completeness is assessed across a fixed panel of traits. However, scope-aware analyses show that sex-stratified variables are not consistently less complete than non-sex-stratified variables, and in some cases are more complete. These findings highlight a central trade-off in comparative life-history analyses: sex-specific measures of traits such as age at sexual maturity are biologically informative, but their inconsistent inclusion across databases complicates broad taxonomic comparisons. In contexts where sex-specific maturity data are sparse or structurally unavailable, age at first reproduction may serve as a valid alternative, as it is more widely reported and unambiguously defined.

Thoughts:

-   is age at maturity even dimorphic enough to justify splitting sex classes?

-   is it better to combine the sex classes for the sake of more data or to rely on the scant sex-specific data points?

-   is AFR a sufficient proxy? or depending on the study AFR may be the preferred measure?

    -   first reproduction typically occurs later than technical sexual maturity & is sensitive to many external factors, should it be treated as a distinct life-history measure or can it serve as a proxy for age at maturity?

-   AnAge cites all other databases in our analysis so how is its completeness and number of species represented lower? more rigorous standard for data?

## other analysis?

```{r}

```

## Sources

1.  Borries, C., Gordon, A. D., & Koenig, A. (2013). Beware of Primate Life History Data: A Plea for Data Standards and a Repository. PLOS ONE, 8(6), e67200. <https://doi.org/10.1371/journal.pone.0067200>

2.  Ernest, S.K.M. (2003), LIFE HISTORY CHARACTERISTICS OF PLACENTAL NONVOLANT MAMMALS. Ecology, 84: 3402-3402. <https://doi.org/10.1890/02-9002>

3.  de Magalhães JP, Costa J. A database of vertebrate longevity records and their relation to other life-history traits. J Evol Biol. 2009 Aug;22(8):1770-4. doi: 10.1111/j.1420-9101.2009.01783.x. Epub 2009 Jun 10. PMID: 19522730.

4.  de Magalhães, J. P., Abidi, Z., Dos Santos, G. A., Avelar, R. A., Barardo, D., Chatsirisupachai, K., Clark, P., De-Souza, E. A., Johnson, E. J., Lopes, I., Novoa, G., Senez, L., Talay, A., Thornton, D., To, P. K. P. (2024) "Human Ageing Genomic Resources: updates on key databases in ageing research." Nucleic Acids Research 52(D1):D900-D908.

5.  IUCN SSC Primate Specialist Group. (2023). *Report of the IUCN Species Survival Commission Primate Specialist Group, Issue 64* (2023). IUCN. <https://iucn.org/sites/default/files/2024-10/2023-iucn-ssc-primate-sg-report_publication.pdf>

6.  Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O’Dell, J., Orme, C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly, C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A., Price, S. A., Rigby, E. A., Rist, J., ... Purvis, A. (2009). PanTHERIA: A species-level database of life history, ecology, and geography of extant and recently extinct mammals. Ecology, 90(9), 2648–2648. <https://doi.org/10.1890/08-1494.1>

7.  Kappeler, P. M., Pereira, M. E., & Schaik, C. P. v. (2003). Primate life histories and socioecology. In P. M. Kappeler & M. E. Pereira (Eds.), *Primate life histories and socioecology* (pp. 1–20). The University of Chicago Press.

8.  Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. and Ernest, S.K.M. (2015), An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles. Ecology, 96: 3109-3109. <https://doi.org/10.1890/15-0846R.1>
