---
title: "An Investigation of Primate Life History Data Quality"
author: "Gentry Miller"
format: 
  html:
    code-overflow: wrap
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: Contents
    df-print: kable
    theme: readable
execute: 
  freeze: auto
editor: visual
---

## if youre reading this im sorry for how much of a mess this doc is right now im mid overhaul after re-applying the permutation test to a more meaningful metric

# Introduction

> “Given this general importance of life history data and with the scientific ideals of transparency and repeatability in mind, one would assume that researchers have long since agreed upon life history definitions and are using reliable and accurate data stored in standardized repositories. However, while various databases exist, standards for data quality and specification are lacking.” (Borries et al. 2013)

Life history theory has provided not just a framework to understand how organisms maximize fitness, but a lexicon of variables that represent a shared currency across organisms. Key variables have emerged to characterize growth (weaning age ), reproduction (age at sexual maturity, first reproduction), and duration of reproduction (lifespan). Despite the widespread use of such measures, researchers face consistent methodological challenges in collating life history data for comparative analyses including access to a substantial range, differences in definitions, discrepancies in data quality, and decision trees on which variables to use and how. 

Here we present a guide that addresses these and other issues for primatologists who want to employ life history data. We review the characteristics of five databases that contain life history data on primates. We also present a frequently advocated approach to summarize variation across traits using dimension reduction analysis, with which we summarize data here. 

The databases assessed here are the PanTHERIA (Jones et al. 2009), AnAge (de Magalhães & Costa 2009), Amniotes (Myrhvold et al. 2015), Ernest (2003), and Kappeler & Pereira (2003) databases on life history statistics.

# Database Analysis

## Preliminaries

### Data Cleaning

```{r}
#| label: data import script
#| echo: false

source("https://raw.githubusercontent.com/bygentry/LHDB-PCA/refs/heads/main/setup.R")
```

#### Filtering Primate Data

Four of the databases under investigation (all but Kappeler & Pereira (2003)) include data for more animals than primates - our focus is on the quality of *primate* life history data, so all other taxa were excluded from our tests.

#### Missing Values and Point Correction

Our .csv of the Kappeler & Pereira database was transcribed by hand from a digital scan of the original paper publication. One value of adult female body mass (`BM.F`) in the Kappeler & Pereira data has a "?", making the object type 'character' instead of 'numeric' . Mixing number and non-number characters would result in a forced NA for that value when converting to type numeric, so the '?' is removed to include the value in completeness tests despite its uncertainty. Additionally, several of the databases use -`999` as the "missing value" - these are converted to `NA` for all fields of type numeric to allow quantitative analysis.

#### Variable Definitions and Selection

Each database records a number of life history statistics, any of which are reported in several of the databases as they are core life history traits (i.e. inter-birth interval, gestation length, etc). See how the variables are defined in each source database [here](https://bygentry.github.io/LHDB-PCA/Variable-Definitions.html). Rather than evaluating the completeness and reliability of every variable across all databases, we focus on a subset of “key life history traits” that characterize primate reproductive life history and are present in all or most of the databases examined. These traits serve as a representative sample of variables commonly used in taxonomic comparisons in life history research:

| Trait                                |
|:-------------------------------------|
| Adult Body Mass (Female)             |
| Adult Body Mass (Male)               |
| Adult Body Mass (Unspecified)        |
| Age at First Reproduction            |
| Age at Sexual Maturity (Female)      |
| Age at Sexual Maturity (Male)        |
| Age at Sexual Maturity (Unspecified) |
| Gestation Length                     |
| Inter-birth Interval                 |
| Lifespan                             |
| Weaning Age                          |

: **Table 1.** "Key Life History Traits" selected for cross-database comparison

```{r}
#| label: building dfs of key vars
#| code-fold: true

sources = as.factor(c("Myrhvold et al. 2015", "de Magalhães & Costa 2009", "Ernest 2003", "Jones et al. 2009", "Kappeler & Pereira 2003"))

## selecting vars for cross-database comparison
amniote_s <- amniote %>% 
  transmute(
    species = paste(genus, species),
    body_mass_u = adult_body_mass_g,
    body_mass_f = female_body_mass_g,
    body_mass_m = male_body_mass_g,
    AFR = female_maturity_d,
    gest_len = gestation_d,
    IBI = inter_litter_or_interbirth_interval_y,
    max_longevity = maximum_longevity_y,
    age_at_maturity_u = NA,
    age_at_maturity_f = female_maturity_d,
    age_at_maturity_m = male_maturity_d,
    weaning_age = weaning_d,
    source = sources[1]
  )

AnAge_s <- AnAge %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = `Body mass (g)`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = NA,
    gest_len = `Gestation/Incubation (days)`,
    IBI = `Inter-litter/Interbirth interval`,
    max_longevity = `Maximum longevity (yrs)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = `Female maturity (days)`,
    age_at_maturity_m = `Male maturity (days)`,
    weaning_age = `Weaning (days)`,
    source = sources[2]
  )

ernest_s <- ernest %>% 
  transmute(
    species = paste(Genus, species),
    body_mass_u = NA,
    body_mass_f = `mass(g)`,
    body_mass_m = NA,
    AFR = `AFR(mo)`,
    gest_len = `gestation(mo)`,
    IBI = `litters/year`,
    max_longevity = `max. life(mo)`,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `weaning(mo)`,
    source = sources[3]
  )

pan_s <- pan %>% 
  transmute(
    species = paste(MSW05_Genus, MSW05_Species),
    body_mass_u = `5-1_AdultBodyMass_g`,
    body_mass_f = NA,
    body_mass_m = NA,
    AFR = `3-1_AgeatFirstBirth_d`,
    gest_len = `9-1_GestationLen_d`,
    IBI = `14-1_InterbirthInterval_d`,
    max_longevity = `17-1_MaxLongevity_m`,
    age_at_maturity_u = `23-1_SexualMaturityAge_d`,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = `25-1_WeaningAge_d`,
    source = sources[4]
  )

kp_s <- kp %>% 
  transmute(
    species = paste(Genus, Species),
    body_mass_u = BM.U,
    body_mass_f = BM.F,
    body_mass_m = NA,
    AFR = AFR,
    gest_len = GL,
    IBI = IBI,
    max_longevity = NA,
    age_at_maturity_u = NA,
    age_at_maturity_f = NA,
    age_at_maturity_m = NA,
    weaning_age = WA,
    source = sources[5]
  )
```

#### Standardizing Units

To make valid comparisons, variables across databases should be in common units. There are several changes to be made to standardize the units to meet these conditions:

1.  Longevity is reported in years

2.  All other time values are reported in days

3.  All mass values are reported in grams

Some of the time values are reported in months - we treat a month as the average length of a month in the Gregorian calendar in days and use these conversion factors throughout unit standardization:

$$ 1 \text{ year } = \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{365.25 \text{ days}}{12 \text{ months}} \approx \frac{30.44 \text{ days}}{1 \text{ month}} \\ $$

The following unit changes are necessary:

**Amniotes (Myrhvold et al. 2015):**

-   Inter-birth Interval (IBI) should be converted from years to days

**AnAge** **(de Magalhães & Costa 2009):**

-   No changes necessary

**Ernest (2003):**

-   Age at First Reproduction (AFR), Gestation Length (gest_len), and Weaning Age (weaning_age) should be converted from months to days

-   IBI should be derived from litters/year

    -   IBI is effectively days/birth (or days/litter) and can be derived from litters/year as such:

        $$ 1 \, \div \frac{x \text{ litters}}{1 \text{ year}} = \frac{x \text{ years}}{1 \text{ litter}} \,\times \, \frac{365.25 \text{ days}}{1 \text{ year}} = \frac{x \text{ days}}{1\text{ litter}}
        $$

-   Maximum Longevity (max_longevity) should be converted from months to years

**PanTHERIA (Jones et al. 2009):**

-   Maximum Longevity should be converted from months to years

**Kappeler & Pereira (2003):**

-   IBI should be converted from months to days

```{r}
#| label: unit conversions
#| code-fold: true

days_per_year <- 365.25
months_per_year <- 12
days_per_month <- days_per_year/months_per_year

amniote_s <- amniote_s %>% 
  mutate(
    IBI = IBI * days_per_year
    # converts years to days
  )

ernest_s <- ernest_s %>% 
  mutate(
    AFR = AFR * days_per_month,
    # converts months to days
    gest_len = gest_len * days_per_month,
    # converts months to days
    IBI = days_per_year/IBI,
    # converts litters/year to IBI (days)
    max_longevity = max_longevity/months_per_year,
    # converts months to years
    weaning_age = weaning_age * days_per_month
    # converts months to days
  )

pan_s <- pan_s %>% 
  mutate(
    max_longevity = max_longevity/months_per_year
    # converts months to years
  )

kp_s <- kp_s %>% 
  mutate(
    IBI = IBI * days_per_month
    # converts months to days
  )
```

### Pre-vis Processing & Stats Test Setup

```{r}
#| label: heat map prep
#| code-fold: true
# Variable % Missing Vals Heat Map Prep

## combine values from key value dfs of each source into a single df "key_vals"
key_vals <- bind_rows(amniote_s, AnAge_s, ernest_s, kp_s, pan_s)

key_long <- key_vals %>%
  pivot_longer(
    cols = -c(species, source),
    names_to = "trait_raw",
    values_to = "value"
  )

trait_labels <- c(
  max_longevity = "Lifespan (Years)",
  IBI = "Interbirth Interval (Days)",
  AFR = "Age at First Reproduction (Days)",
  gest_len = "Gestation Length (Days)",
  weaning_age = "Weaning Age (Days)",
  body_mass_u = "Adult Mass - Unspecified (g)",
  body_mass_m = "Adult Mass - Male (g)",
  body_mass_f = "Adult Mass - Female (g)",
  age_at_maturity_u = "Sexual Maturity - Unspecified (Days)",
  age_at_maturity_m = "Sexual Maturity - Male (Days)",
  age_at_maturity_f = "Sexual Maturity - Female (Days)"
)

scope_df <- key_long %>%
  group_by(source, trait_raw) %>%
  summarise(
    n_total = n(),
    n_nonmissing = sum(!is.na(value)),
    prop_missing_panel = mean(is.na(value)),
    .groups = "drop"
  ) %>%
  mutate(
    in_scope = (n_nonmissing / n_total) >= 0.01,
    prop_missing_within_scope = if_else(in_scope, prop_missing_panel, NA_real_),
    trait = recode(trait_raw, !!!trait_labels)
  )

scope_summary <- scope_df %>%
  group_by(source) %>%
  summarise(
    scope_coverage = mean(in_scope),                              
    panel_completeness = 1 - mean(prop_missing_panel),            
    within_scope_completeness = 1 - mean(prop_missing_within_scope, na.rm = TRUE),
    .groups = "drop"
  )

## this calculates the proportion of missing values for each life history trait
## and stores them in a "missigness" df that stores the trait, the source, 
## and the % missingness in 3 steps
## (1) pivots columns except for species and source, 
##     assigns variable names to trait and values to value 
## (2) groups traits sourcewise
## (3) calculates the proportion of NAs to valid entries for each trait
##     in each source and stores them alongside the source and trait name
missingness <- key_vals %>%
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")

missingness_no_sex <- key_vals %>%
  select(-any_of(c("body_mass_u", "body_mass_f", "body_mass_m", "age_at_maturity_u", "age_at_maturity_f", "age_at_maturity_m"))) %>% 
  pivot_longer(cols = -c(species, source), 
               names_to = "trait", 
               values_to = "value") %>%
  group_by(source, trait) %>% 
  summarize(prop_missing = mean(is.na(value)), .groups = "drop")


missingness <- missingness %>%
  mutate(trait = dplyr::recode(trait, !!!trait_labels, .default = trait))

missingness_no_sex <- missingness_no_sex %>%
  mutate(trait = dplyr::recode(trait, !!!trait_labels, .default = trait))

## extracts the number of distinct species represented by each source
## then concatenates a label of the citation and number of primate species
dataset_labels <- key_vals %>%
  group_by(source) %>%
  summarize(n = n_distinct(species), .groups = "drop") %>%
  mutate(label = paste0(source, "\n(n = ", n, ")")) %>% 
  { setNames(.$label, .$source) }
```

```{r}
#| label: permutation test
#| code-fold: true
# One-tailed Permutation Test

# create list of sex dependent variable names for checking sex differentiation
sex_dep_vars <- c("Sexual Maturity - Female (Days)", "Sexual Maturity - Male (Days)", "Sexual Maturity - Unspecified (Days)", "Adult Mass - Female (g)", "Adult Mass - Male (g)", "Adult Mass - Unspecified (g)")

# create a new df for the permutation test
## completeness - the inverse proportion of missing values
## sexed - a boolean representing a trait is stratified by sex
perm_df <- scope_df %>% 
  filter(in_scope) %>% 
  mutate(
    completeness = 1 - prop_missing_panel,        # inverse proportion of missingness for each trait
    sexed = trait %in% sex_dep_vars         # boolean: TRUE if variable is sex stratified, else FALSE 
  )

# creates a function for the permutation test
perm_test <- function(df, n = 10000, seed = 1){
  set.seed(seed)

  x <- df$completeness
  sexed <- df$sexed

  n_sexed <- sum(sexed)
  n_non <- sum(!sexed)

  if(n_sexed == 0 || n_non == 0){
    return(tibble(
      mean_sexed = NA_real_,
      mean_non = NA_real_,
      obs_diff = NA_real_,
      p_value = NA_real_,
      n_sexed = n_sexed,
      n_non = n_non
    ))
  }

  mu_sexed <- mean(x[sexed])
  mu_non <- mean(x[!sexed])
  obs <- mu_non - mu_sexed

  perm_stats <- replicate(n, {
    perm <- sample(sexed)
    mean(x[!perm]) - mean(x[perm])
  })

  tibble(
    mean_sexed = mu_sexed,
    mean_non = mu_non,
    obs_diff = obs,
    p_value = mean(perm_stats >= obs), 
    n_sexed = n_sexed,
    n_non = n_non
  )
}

```

```{r}
#| label: completeness bar chart prep
#| code-fold: true
# Completeness Bar Chart Prep

## makes a new df using prop_missing values from 
## missingness df and averages them by source, then stores
## in new df mean_missing
mean_missing <- missingness %>% 
  group_by(source) %>% 
  summarise(
    mean_missingness = mean(prop_missing)
    )

## takes the inverse proportion of the mean missingness
## for each database to give mean completeness
mean_complete <- mean_missing %>% 
  mutate(mean_completeness = 1 - mean_missingness)

## removing sexed variables
mean_missing_no_sex <- missingness_no_sex %>% 
  group_by(source) %>% 
  summarise(
    mean_missingness_no_sex = mean(prop_missing)
  )

## takes the inverse proportion of the mean missingness
## for each database to give mean completeness
mean_complete_no_sex <- mean_missing_no_sex %>% 
  mutate(mean_completeness_no_sex = 1 - mean_missingness_no_sex)

composite_completeness <- mean_complete %>% 
  inner_join(mean_complete_no_sex, by = "source") %>% 
  mutate(
    cap = pmax(0, mean_completeness_no_sex - mean_completeness)
  )
```

## Proportions of Missing Values

One of the primary issues with primate life history data is how few species are represented. None of the databases inspected in this study represent more than 70% of primate species, highlighting the ever-present need for more data.

| Database | N species | Proportion of Primates Represented\* |
|----|----|----|
| Amniotes (Myhrvold et al. 2015) | 365 | .68 |
| AnAge (de Magalhães & Costa 2009) | 174 | .33 |
| Ernest (2003) | 156 | .29 |
| PanTHERIA (Jones et al. 2009) | 376 | .70 |
| Kappeler & Pereira (2003) | 310 | .58 |

: **Table 2.** Proportions of primate species represented by each database. \*Calculated as N/535, the number of species recognized by the IUCN / SSC Primate Specialist Group (2023)

Of the species we have data for, many are lacking at least one important life history statistic. Heat map analysis of the proportion of missing values to data points for each key life history trait reveal several patterns in the short-comings of the data we have available:

#### Visualization & Interpretation

```{r}
#| label: heatmap visualization 1
#| fig-cap: "**Figure 1.** Heat map of missing values for key life history traits across all five databases. Low heat (yellow) signifies few missing values while high heat (red) signifies many missing values. Grey indicates that the trait was not recorded in a particular source."
#| echo: false

ggplot(scope_df, aes(x = source, y = trait)) +
  geom_tile(aes(fill = prop_missing_within_scope), color = "snow") +
  geom_tile(
    data = dplyr::filter(scope_df, !in_scope),
    fill = "grey85",
    color = "snow"
  ) +
  scale_fill_gradientn(
    colors = rev(colorspace::heat_hcl(100)),
    limits = c(0, 1),
    breaks = seq(0, 1, by = 0.25),
    labels = scales::label_percent(accuracy = 1),
    name = "% Missing\n(within scope)",
    na.value = "grey85"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Proportions of Missing Values for Key Traits",
    x = "Dataset",
    y = "Trait"
  )

```

notes:

-   sex-reliant vars are most unreported

-   age at maturity vars for females are more complete than male when both sexes are reported

-   weaning age, IBI, and gestation length are the most complete across all data sets (lifespan is a close 4th but K&P don't report any values for it - same with AnAge and AFR )

-   **all databases have at least one unreported variable of our key variable selection**

Of our key life history traits, most databases only record data for around 60% of them (mostly just **not** the sex-stratified versions of body mass and age at maturity). When testing the mean completeness of the databases using all of the key traits versus only the ones they record, their completeness scores obviously improved. This may appear to call into question our selection of traits, but really the trait scope is under represented because the missing traits are the same measurement for different sex classes.

```{r}
scope_summary %>%
  mutate(
    scope_coverage = scales::percent(scope_coverage, accuracy = 1),
    panel_completeness = round(panel_completeness, 3),
    within_scope_completeness = round(within_scope_completeness, 3)
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    source = "Database",
    scope_coverage = "Trait scope coverage",
    within_scope_completeness = "Completeness (within scope)",
    panel_completeness = "Completeness (full panel)"
  ) %>%
  gt::tab_caption(gt::md("**Table 3.** Scope coverage vs completeness (panel vs within-scope)."))
```

I ran a one-tailed permutation test to compare the mean completeness of the databases using all of our key LH variables and the mean completeness when the sex-stratified variables are excluded:

```{r}
#| label: permutation test results
#| echo: false
perm_res <- perm_df %>%
  group_by(source) %>%
  group_modify(~ perm_test(.x)) %>%
  ungroup()

perm_res %>%
  mutate(
    mean_sexed = round(mean_sexed, 3),
    mean_non = round(mean_non, 3),
    obs_diff = round(obs_diff, 3),
    p_value = signif(p_value, 3)
  ) %>%
  gt() %>%
  cols_label(
    source = "Database",
    mean_non = "Non-sex-stratified traits",
    mean_sexed = "Sex-stratified traits",
    obs_diff = "Δ completeness (non − sexed)",
    p_value = "p (one-tailed)",
    n_non = "k (non)",
    n_sexed = "k (sexed)"
  ) %>%
  tab_spanner(
    label = "Mean completeness (within scope)",
    columns = c(mean_non, mean_sexed)
  ) %>%
  tab_caption(
    md("**Table 4.** Scope-aware permutation test comparing completeness of non-sex-stratified vs sex-stratified traits within each database’s in-scope trait set (n = 10,000 label permutations). One-tailed p tests whether non-sex-stratified traits are more complete.")
  )

```

Removing the sex-dependent variables improves completeness across the board, but the difference from a null distribution was only "significant" at p \< 0.05 for Ernest (2003) (p = 0.013). However, because Ernest's database doesn't report these variables to begin with, we are essentially just making the comparison more fair by not including unrecorded data. A similar effect takes place in the Amniote (Myrhvold et al. 2015) and AnAge (de Magalhães & Costa 2009) databases, which both had suggestive increases in mean completeness at p \< 0.05 (0.052 and 0.069, respectively) when not testing for unreported sex-dependent variables. The variables that were the most complete across databases were also all non-stratified: weaning age, gestation length, and IBI. Lifespan and age at first reproduction have moderate completeness across all the databases except for AnAge, which doesn't report AFR, and the K&P dataset, which doesn't report lifespan.

```{r}
#| label: layered bar chart
#| fig-cap: "**Figure 2.** Mean completeness of each database using all key variables (solid color) and only non-sex-stratified key variables (subset, translucent color). Numeric p-values indicate results of one-tailed permutation tests comparing the observed difference in mean completeness to the null distribution generated by random subsets of equal size (n = 10,000)"
#| echo: false

#plot_pvals <- perm_res %>%
#  mutate(
#    label = paste0("p = ", format.pval(p_value, digits = 2, eps = 0.001)),
#    y_pos = pmin(mean_completeness_no_sex + 0.04, 0.98)
#  )

#ggplot(composite_completeness, aes(x = source)) +
#  geom_col(
#    aes(y = mean_completeness_no_sex, fill = source, color = source),
#    alpha = 0.4,
#    width = 0.75,
#    linewidth = 0.5,
#    show.legend = FALSE
#  ) +
#  geom_col(
#    aes(y = mean_completeness, fill = source),
#    width = 0.75
#  ) +
#  geom_text(
#    data = plot_pvals,
#    aes(x = source, y = y_pos, label = label),
#    inherit.aes = FALSE,
#    size = 4
#  ) +
#  coord_cartesian(ylim = c(0, 1)) +
#  scale_fill_viridis_d(option = "D", end = 0.9) +
#  scale_color_viridis_d(option = "D", end = 0.9, guide = "none") +
#  labs(
#    x = "Source Database",
#    y = "Mean Completeness",
#    title = "Mean Completeness of Sexless vs All Key Traits"
#  ) +
#  theme_bw() +
#  theme(
#    legend.position = "none",
#    axis.text.x = element_text(angle = 33, hjust = 1)
#  )
```

The PanTHERIA (blue) and Kappeler & Pereira (green) databases both had poor completeness scores across all variables. For select species with high representation, like humans or chimpanzees, the Kappeler & Pereira set seems like a good inclusion but it doesn't have much to offer in wide taxonomic comparisons. Ernest (dark blue) has the second highest mean completeness when only considering sex-independent variables but falls short in the number of primate species represented (only 156). Similarly, the AnAge database (purple) demonstrates high completeness for key variables but only contains data for 174 species of primates. PanTHERIA, while only having a mean completeness of 32% with sex-independent variables, boasts the greatest representation of primates with 376 species. In both representation and completeness, the Amniote database (yellow) is arguably the best. While only representing 365 primate species, its mean completeness is 61% for all variables and 76% for sex-independent variables.

NOTE: AnAge cites all other databases in our analysis so how is its completeness and number of species represented lower? more rigorous standard for data?

Considering the limited effect of including sex-dependent variables on the completeness of the PanTHERIA (Jones et al. 2009) and Kappeler & Pereira (2003) databases, and the significant improvement to completeness of the three databases which don't report sex-dependent variables for adult mass and age at maturity, it is my opinion that including these variables only serves to make comparisons between databases less meaningful and more complicated computationally. I remember us debating including body mass to begin with, but couldn't find in our notes whether we came to a conclusion. Age at maturity is irrefutably an important life history statistic, though, so the scarcity of sex-specific data can pose a problem to researchers. One alternative is age at first reproduction, which is a more definitive way of knowing that an individual has reached maturity but doesn't necessarily correlate to when they *reached* maturity, because first reproduction tends to occur later than when it is technically possible.

Thoughts:

-   is age at maturity even dimorphic enough to justify splitting sex classes?

-   is it better to combine the sex classes for the sake of more data or to rely on the scant sex-specific data points?

-   is AFR a sufficient proxy? or depending on the study AFR may be the preferred measure?

## other analysis?

```{r}

```

## Sources

1.  Borries, C., Gordon, A. D., & Koenig, A. (2013). Beware of Primate Life History Data: A Plea for Data Standards and a Repository. PLOS ONE, 8(6), e67200. <https://doi.org/10.1371/journal.pone.0067200>

2.  Ernest, S.K.M. (2003), LIFE HISTORY CHARACTERISTICS OF PLACENTAL NONVOLANT MAMMALS. Ecology, 84: 3402-3402. <https://doi.org/10.1890/02-9002>

3.  de Magalhães JP, Costa J. A database of vertebrate longevity records and their relation to other life-history traits. J Evol Biol. 2009 Aug;22(8):1770-4. doi: 10.1111/j.1420-9101.2009.01783.x. Epub 2009 Jun 10. PMID: 19522730.

4.  de Magalhães, J. P., Abidi, Z., Dos Santos, G. A., Avelar, R. A., Barardo, D., Chatsirisupachai, K., Clark, P., De-Souza, E. A., Johnson, E. J., Lopes, I., Novoa, G., Senez, L., Talay, A., Thornton, D., To, P. K. P. (2024) "Human Ageing Genomic Resources: updates on key databases in ageing research." Nucleic Acids Research 52(D1):D900-D908.

5.  IUCN SSC Primate Specialist Group. (2023). *Report of the IUCN Species Survival Commission Primate Specialist Group, Issue 64* (2023). IUCN. <https://iucn.org/sites/default/files/2024-10/2023-iucn-ssc-primate-sg-report_publication.pdf>

6.  Jones, K. E., Bielby, J., Cardillo, M., Fritz, S. A., O’Dell, J., Orme, C. D. L., Safi, K., Sechrest, W., Boakes, E. H., Carbone, C., Connolly, C., Cutts, M. J., Foster, J. K., Grenyer, R., Habib, M., Plaster, C. A., Price, S. A., Rigby, E. A., Rist, J., ... Purvis, A. (2009). PanTHERIA: A species-level database of life history, ecology, and geography of extant and recently extinct mammals. Ecology, 90(9), 2648–2648. <https://doi.org/10.1890/08-1494.1>

7.  Kappeler, P. M., Pereira, M. E., & Schaik, C. P. v. (2003). Primate life histories and socioecology. In P. M. Kappeler & M. E. Pereira (Eds.), *Primate life histories and socioecology* (pp. 1–20). The University of Chicago Press.

8.  Myhrvold, N.P., Baldridge, E., Chan, B., Sivam, D., Freeman, D.L. and Ernest, S.K.M. (2015), An amniote life-history database to perform comparative analyses with birds, mammals, and reptiles. Ecology, 96: 3109-3109. <https://doi.org/10.1890/15-0846R.1>
